---
title: "Data in R"
subtitle: "Module I · Day 1 · 3 hours"
date: "2026-04-23"
execute:
  echo: true
  warning: false
  message: false
---

## Outline

::: {.incremental}
1. R, RStudio, Quarto, tidyverse — what each is for.
2. R as a calculator · expressions, assignment, functions.
3. **Atomic types** · numeric, character, logical, factor, date.
4. **Vectors** · creation, arithmetic, subsetting.
5. **Data frames and tibbles** · the rectangular data model.
6. **Importing tourism data** · CSV, Excel, SPSS, Eurostat API.
:::

# 1 · R, RStudio, Quarto, tidyverse {background-color="#0067a2"}

## Four pieces, one stack

|  | What it is | You write… |
|---|---|---|
| **R** | The language and engine. | Code that computes. |
| **RStudio / Positron** | The IDE where you work. | Nothing R-specific — it's the editor. |
| **Quarto** | The publishing system. | Reports, slides, web pages, books. |
| **tidyverse** | A family of R packages with consistent design. | `library(tidyverse)` at the top of every script. |

Install R from <https://cloud.r-project.org>, RStudio from <https://posit.co/download/rstudio-desktop/>.

## The RStudio layout

```
┌───────────────────────────────┬──────────────────────────┐
│                               │                          │
│   Editor                      │   Environment / History  │
│   (your .R or .qmd files)     │   (objects currently     │
│                               │    in memory)            │
│                               │                          │
├───────────────────────────────┼──────────────────────────┤
│                               │                          │
│   Console                     │   Files / Plots /        │
│   (live R session)            │   Packages / Help        │
│                               │                          │
└───────────────────────────────┴──────────────────────────┘
```

- Write in the **editor**, send a line with **Ctrl + Enter** (**Cmd + Enter** on Mac).
- The **console** shows the output.
- **Files** pane is your file browser; **Git** pane appears when the folder is a repo.

## Projects · always work inside an `.Rproj`

- Double-click `quantitative-methods-master-tides.Rproj` to open the project.
- Working directory is fixed to the project root → paths are portable.
- Packages, history and environment are isolated per project.

::: {.callout-tip}
**Never** run `setwd("C:/Users/yourname/...")`. If you see that in any
tutorial, close the tab.
:::

## Packages · extending R

Base R is a small kernel. Everything useful — tidyverse, plotting,
model fitting, reading Excel, hitting the Eurostat API — ships as a
**package** that you install separately from
[CRAN](https://cran.r-project.org), the canonical mirror.

```{r}
#| eval: false
# install once per laptop
install.packages("tidyverse")
install.packages(c("readxl", "haven", "here", "eurostat"))

# load at the top of every script that uses them
library(tidyverse)
library(here)
```

::: {.callout-tip}
`install.packages()` downloads and compiles — slow, but only once.
`library()` is fast and goes at the top of every script. One single
`library(tidyverse)` pulls in eight core packages (`dplyr`,
`ggplot2`, `tibble`, `readr`, `tidyr`, `purrr`, `stringr`,
`forcats`) — the workhorses for the rest of the course.
:::

# 2 · R as a calculator {background-color="#0067a2"}

## First expressions

```{r}
2 + 2
17 %% 5        # modulo
2 ^ 10         # power
sqrt(81)
log(exp(1))    # natural log of e
```

R evaluates each line top-to-bottom and prints the result. Nothing else happens.

## Assignment

```{r}
visitors <- 1250          # store a value in a name
nights   <- 4.2
spending <- 85

total <- visitors * nights * spending
total
```

Use `<-` (`Alt` + `-` in RStudio inserts it) for assignment. Names can
contain letters, digits, `_` and `.`, but must **start with a letter**.

## Functions

Functions take inputs and return an output:

```{r}
round(4.2, digits = 0)
mean(c(2, 5, 7, 10, 14))
seq(from = 2024, to = 2026, by = 1)
```

::: {.callout-tip}
Type `mean(` and press **F1** in RStudio to open the help page for the
function. `?mean` does the same from the console.
:::

## Getting help

```{r}
#| eval: false
?mean              # help for one function
??"linear model"   # search across packages
example(lm)        # run the documented examples
```

Sites you will use a lot:

- <https://stat.ethz.ch/R-manual/> — canonical reference.
- <https://r4ds.hadley.nz/> — R for Data Science, 2nd edition (free online).
- Stack Overflow, tagged `[r]`.

# 3 · Atomic types {background-color="#0067a2"}

## The five types you meet every day

```{r}
typeof(3.14)              # "double"   — real numbers
typeof(1L)                # "integer"  — note the L suffix
typeof("Gran Canaria")    # "character"
typeof(TRUE)              # "logical"
class(factor(c("a","b","a")))   # "factor"
class(Sys.Date())         # "Date"
```

We lump *numeric* (double + integer) together in practice.

## Numeric

```{r}
price    <- 82.50            # double
stars    <- 4L               # integer
is.numeric(price)
is.numeric(stars)

# be careful with floating-point comparisons
0.1 + 0.2 == 0.3             # FALSE!
all.equal(0.1 + 0.2, 0.3)    # TRUE
```

## Character

```{r}
island <- "Lanzarote"
nchar(island)
toupper(island)

# paste strings together
paste("Hotel", island, sep = " · ")

# case matters
"Lanzarote" == "lanzarote"
```

## Logical

```{r}
is_open <- TRUE
T == TRUE                # TRUE — but don't use T, it can be overwritten
sum(c(TRUE, FALSE, TRUE, TRUE))   # 3 — logicals coerce to 0/1
```

Operators return logicals:

```{r}
82.50 > 70
"Lanzarote" %in% c("Tenerife", "Lanzarote", "La Palma")
```

## Factor · categories with known levels

```{r}
island <- factor(
  c("Gran Canaria", "Tenerife", "Gran Canaria", "Lanzarote"),
  levels = c("Gran Canaria", "Tenerife", "Lanzarote",
             "Fuerteventura", "La Palma", "La Gomera", "El Hierro")
)

levels(island)
table(island)
```

Factors matter when the order of categories is meaningful (e.g. *low*,
*medium*, *high*) or when all levels should appear in a table even if
some have zero observations.

## Date and time

```{r}
today <- Sys.Date()
today
class(today)

# arithmetic on dates
today + 30
as.Date("2026-06-08") - today     # days to the first exam

format(today, "%A, %d %B %Y")     # pretty-print
```

Use the `lubridate` package (inside the tidyverse) for anything more
serious — we'll cover it in *Data wrangling*.

## Coercion gotchas

```{r}
c(1, 2, "three")      # everything becomes character
c(1, 2, TRUE)         # TRUE becomes 1 — all numeric
c(1L, 2L, 3.14)       # integer → double
```

R's rule: when mixing types in one vector, coerce **up** to the most
flexible type (`character` > `double` > `integer` > `logical`).

# 4 · Vectors {background-color="#0067a2"}

## One type per vector

An atomic vector is **homogeneous**: every element must be of the
**same type** — all `double`, or all `integer`, or all `character`,
or all `logical`, etc.

```{r}
typeof(c(1.5, 2.7, 3.0))          # "double"
typeof(c(1L, 2L, 3L))             # "integer"
typeof(c("GC", "TF", "LZ"))       # "character"
typeof(c(TRUE, FALSE, TRUE))      # "logical"
```

::: {.callout-important}
Mixing types is not forbidden — R silently coerces everything to one
common type (see *Coercion gotchas* below), which is almost never
what you want. If you need several types in one object, use a
**data frame**: one column per type.
:::

## Creating vectors

```{r}
# combine — the main way to build a vector
nights  <- c(2, 5, 7, 10, 14)
islands <- c("Gran Canaria", "Tenerife", "Lanzarote",
             "Fuerteventura", "La Palma")

# sequences
1:5
seq(from = 0, to = 1, by = 0.25)
rep("GC", times = 3)

length(nights)
```

## Vectorised arithmetic

```{r}
price   <- c(82, 95, 110, 100, 78)
revenue <- nights * price                 # element-wise — no loops
revenue

mean(revenue)
sum(revenue)
range(revenue)       # min and max
```

This is the point of R. *Whenever* you are tempted to write a `for`
loop, ask yourself whether vectorisation solves it.

## Subsetting vectors

```{r}
nights
nights[1]             # first element — R is 1-indexed
nights[c(1, 3, 5)]    # several positions
nights[-1]            # everything EXCEPT the first
nights[nights > 5]    # logical subsetting — most useful
```

The logical form (`x[condition]`) is the one you will use 95 % of the time.

## Named vectors

```{r}
price <- c(
  "Gran Canaria"  = 82,
  "Tenerife"      = 95,
  "Lanzarote"     = 110,
  "Fuerteventura" = 100,
  "La Palma"      = 78
)

price["Tenerife"]
price[price > 90]
```

Named vectors are a stepping stone to data frames.

# 5 · Data frames and tibbles {background-color="#0067a2"}

## The rectangular data model

```{r}
library(tibble)

hotels <- tibble(
  island = c("Gran Canaria", "Tenerife", "Lanzarote",
             "Fuerteventura", "La Palma"),
  stars  = c(4L, 5L, 4L, 3L, 3L),
  price  = c(82, 95, 110, 100, 78),
  nights = c(12.5, 18.3, 9.8, 11.2, 6.4)
)

hotels
```

- Rows = observations (one hotel per row).
- Columns = variables (each has one type).
- Every column has the **same length**.

## Inspecting a data frame

```{r}
dim(hotels)      # rows, columns
nrow(hotels)
ncol(hotels)
names(hotels)
```

```{r}
#| eval: false
head(hotels)     # first 6 rows
tail(hotels)     # last 6 rows
glimpse(hotels)  # tidyverse — compact overview
summary(hotels)  # base R — stats per column
View(hotels)     # spreadsheet view (RStudio)
```

## Selecting columns

```{r}
hotels$price           # dollar notation — a single column as a vector
hotels[["price"]]      # equivalent, bracket form
hotels[, "price"]      # still a one-column tibble

mean(hotels$price)
```

## Adding and modifying columns

```{r}
hotels$revenue <- hotels$price * hotels$nights
hotels
```

Later, with `dplyr`, we'll write this as:

```{r}
#| eval: false
library(dplyr)
hotels <- hotels |> mutate(revenue = price * nights)
```

## Filtering rows

```{r}
# base R
hotels[hotels$stars >= 4, ]

# dplyr preview
library(dplyr)
hotels |> filter(stars >= 4)
```

Pick one style and stick to it. We'll standardise on **dplyr** from
*Data wrangling* onwards.

# 6 · Importing tourism data {background-color="#0067a2"}

## The usual suspects

| Format | Package | Function |
|---|---|---|
| CSV | `readr` | `read_csv()` |
| Excel (`.xlsx`, `.xls`) | `readxl` | `read_excel()` |
| SPSS (`.sav`) | `haven` | `read_sav()` |
| Stata (`.dta`) | `haven` | `read_dta()` |
| JSON | `jsonlite` | `read_json()` |
| Eurostat | `eurostat` | `get_eurostat()` |

All of these are part of — or play nicely with — the tidyverse.

## CSV with `readr::read_csv`

```{r}
#| eval: false
library(readr)
library(here)

occupancy <- read_csv(
  here("datasets", "raw", "istac-nights-by-island.csv"),
  locale = locale(decimal_mark = ",", grouping_mark = ".")
)

glimpse(occupancy)
```

::: {.callout-tip}
Always go through `here::here()`. Your script will work on the lab
machines, on your laptop, and on my laptop with zero changes.
:::

## Excel with `readxl`

```{r}
#| eval: false
library(readxl)

capacity <- read_excel(
  here("datasets", "raw", "istac-hotel-capacity.xlsx"),
  sheet = "2024",
  skip  = 3           # skip the first three header rows
)

glimpse(capacity)
```

Excel files often have merged cells, titles and footers. Use `skip` and
`range = "B5:K120"` to carve out the actual data table.

## SPSS and Stata with `haven`

```{r}
#| eval: false
library(haven)

survey <- read_sav(
  here("datasets", "raw", "tourism-survey.sav")
)

# haven preserves SPSS variable and value labels
attributes(survey$origin_country)
```

Use `as_factor(survey)` to convert labelled numeric variables into R
factors in one go.

## Eurostat · live from the API

```{r}
#| eval: false
library(eurostat)
library(dplyr)

nights <- get_eurostat("tour_occ_nim", time_format = "date") |>
  filter(unit    == "NR",     # number of nights
         c_resid == "TOTAL",  # all residency statuses
         nace_r2 == "I551")   # hotels only

nights
```

The `eurostat` package caches downloads locally, so the slow step only
happens the first time.

## Reproducible paths with `here::here()`

```{r}
#| eval: false
library(here)

here()
# [1] "C:/Users/you/.../quantitative-methods-master-tides"

here("datasets", "raw", "eurostat-nights.csv")
# [1] "C:/Users/you/.../datasets/raw/eurostat-nights.csv"
```

`here()` always resolves relative to the project root (the folder with
the `.Rproj`), regardless of which subfolder your `.R` or `.qmd` file
lives in.

## A first tourism pipeline {.smaller}

```{r}
library(dplyr)
library(tibble)

hotels <- tribble(
  ~island,          ~month,  ~nights,  ~beds,
  "Gran Canaria",   "2024-06",  142500, 165000,
  "Gran Canaria",   "2024-07",  168300, 165000,
  "Tenerife",       "2024-06",  198200, 220000,
  "Tenerife",       "2024-07",  231100, 220000,
  "Lanzarote",      "2024-06",   88100, 105000,
  "Lanzarote",      "2024-07",   98400, 105000
)

hotels |>
  mutate(occupancy = nights / beds) |>
  group_by(island) |>
  summarise(mean_occupancy = mean(occupancy), .groups = "drop") |>
  arrange(desc(mean_occupancy))
```

This is essentially every report in tourism statistics: *load · clean ·
group · summarise*. Tomorrow we unpack each verb.

## Recap

::: {.incremental}
- R objects climb a ladder: **atomic vectors → named vectors → data frames**.
- Learn the five types (`numeric`, `character`, `logical`, `factor`, `Date`) and you understand 90 % of what R prints.
- Arithmetic on vectors is element-wise — avoid loops.
- `tibble` / `data.frame` is the 99 %-of-the-time workhorse.
- Import with the package that matches the file: `readr`, `readxl`, `haven`, `eurostat`.
- Always use `here::here()` for paths.
:::

# Next up {background-color="#0067a2"}

Short break, then one hour on **Git & GitHub** to set up your submission workflow.

After that, the Day 1 exercise: open `exercises/day1/exercise-template.R`,
load one ISTAC CSV and one Eurostat dataset, and inspect them with
`glimpse()`, `summary()` and `head()`.
