---
title: "Data in R"
subtitle: "Topic 1 · Day 1 · 3 hours"
date: last-modified
---

## Outline

::: {.incremental}
- R, RStudio, and the tidyverse: what each is for.
- R as a calculator: expressions, assignment, functions.
- Atomic types: numeric, character, logical, factor, date.
- Vectors and their arithmetic.
- Data frames and tibbles: the rectangular data model.
- Importing tourism data: CSV, Excel, SPSS/Stata.
:::

## R and RStudio

- **R** is the language and the engine.
- **RStudio / Positron** is the IDE — four panes: editor, console, environment, files.
- **Quarto** is how we write slides, reports, dashboards in R.
- **tidyverse** is a family of packages with consistent design — we use it throughout.

## First commands

```{r}
#| echo: true
2 + 2
x <- c(1, 2, 3, 5, 8)
mean(x)
sqrt(x)
```

## Atomic types

```{r}
#| echo: true
typeof(3.14)           # "double"
typeof("Canarias")     # "character"
typeof(TRUE)           # "logical"
class(factor(c("a","b","a")))   # "factor"
class(Sys.Date())      # "Date"
```

## Vectors

- Every value in R is a vector. Scalars are vectors of length 1.
- Arithmetic is vectorised — no loops needed for element-wise ops.

```{r}
#| echo: true
nights  <- c(2, 5, 7, 10, 14)
price   <- c(65, 85, 90, 120, 140)
revenue <- nights * price
revenue
```

## Data frames and tibbles

```{r}
#| echo: true
library(tibble)
hotels <- tibble(
  island  = c("Gran Canaria", "Tenerife", "Lanzarote"),
  nights  = c(12.5, 18.3, 9.8),
  price   = c(82, 95, 110)
)
hotels
```

## Importing data · the usual suspects

| Format | Package | Function |
|---|---|---|
| CSV | `readr` | `read_csv()` |
| Excel (.xlsx, .xls) | `readxl` | `read_excel()` |
| SPSS (.sav) | `haven` | `read_sav()` |
| Stata (.dta) | `haven` | `read_dta()` |
| JSON | `jsonlite` | `read_json()` |
| Eurostat | `eurostat` | `get_eurostat()` |

## Tourism example · ISTAC hotel occupancy

```{r}
#| echo: true
#| eval: false
library(readr)
library(here)

occupancy <- read_csv(
  here("datasets", "raw", "istac-hotel-occupancy.csv"),
  locale = locale(decimal_mark = ",", grouping_mark = ".")
)

head(occupancy)
```

::: {.callout-tip}
Always use `here::here()` — never hard-code absolute paths.
:::

## Recap

- R objects: **atomic vectors** → **data frames** (the 99 % case).
- Import with the matching package: `readr`, `readxl`, `haven`, `eurostat`.
- `here::here()` for portable paths.

# Hands-on {background-color="#0067a2"}

Load one ISTAC tourism CSV and one Eurostat dataset, inspect them with
`glimpse()`, `summary()`, and `head()`. Commit & push.
