---
title: "Git & GitHub essentials"
subtitle: "Module I · Day 1 · 1 hour"
date: "2026-04-23"
---

## Why version control?

::: {.incremental}
- Keep a **full history** of every change — who, what, when, why.
- Collaborate without emailing `analysis_final_v3_reallyfinal.R`.
- **Revert safely** when an experiment breaks your code.
- Make your work **reproducible** and publicly auditable.
:::

::: {.callout-important}
In this course **every exercise you submit lives on GitHub**. The
workflow has to be working for you by the end of this hour.
:::

## Git vs GitHub

|  | **Git** | **GitHub** |
|---|---|---|
| What it is | A tool that tracks changes | A web service that hosts Git projects |
| Runs | Locally, on your laptop | On github.com |
| Needed for | Any version control at all | Sharing, collaboration, PRs |
| Cost | Free, open source | Free for public repos and education |

You need **both**: Git installed locally, a GitHub account online.

## The three Git areas

```
┌─────────────┐   git add   ┌─────────────┐  git commit  ┌────────────────┐
│   Working   │ ──────────▶ │   Staging   │ ───────────▶ │  Local repo    │
│  directory  │             │   (index)   │              │  (your laptop) │
└─────────────┘             └─────────────┘              └────────────────┘
                                                                  │
                                                         git push │
                                                                  ▼
                                                         ┌────────────────┐
                                                         │  GitHub remote │
                                                         └────────────────┘
```

In this course you drive every arrow from **RStudio's Git pane**, not
from a terminal.

# Setup (one-off) {background-color="#0067a2"}

## Step 1 · Install Git

- **macOS** · `brew install git`, or install the Xcode Command Line Tools.
- **Windows** · download [Git for Windows](https://git-scm.com/download/win) and accept the defaults.
- **Linux** · `sudo apt install git` (Debian/Ubuntu) or `sudo dnf install git` (Fedora).

After install, restart RStudio. It detects Git automatically.

## Step 2 · Set your identity — no terminal

From the R console:

```{r}
#| eval: false
usethis::use_git_config(
  user.name  = "Your Name",
  user.email = "you@ulpgc.es"
)
```

Every commit you ever make will be signed with this name and email.
Use the same email you will register with on GitHub (next step).

## Step 3 · GitHub account + PAT {.smaller}

**Create a GitHub account** at <https://github.com/signup> with your
ULPGC email. Apply for the free
[Student Developer Pack](https://education.github.com/pack) — private
repos, Copilot and more.

**Create a Personal Access Token (PAT)** from R:

```{r}
#| eval: false
usethis::create_github_token(
  scopes      = c("repo", "user", "workflow"),
  description = "R on my laptop · TIDES course",
  expiration  = 90                 # 90 days covers the whole course
)
```

The browser opens on GitHub. Leave the fields as prefilled, confirm
**90 days**, click **Generate token**. **Copy the token immediately** —
GitHub never shows it again.

## Step 3 · Save the token

Back in R, paste the token into the credential store:

```{r}
#| eval: false
gitcreds::gitcreds_set()
```

## Verify everything works

```{r}
#| eval: false
usethis::git_sitrep()
```

All rows should show ✅ green. If anything is red, re-run the step
the report points to.

## Core vocabulary {.smaller}

::: {.incremental}
- **Repository (repo)** · a project folder tracked by Git.
- **Commit** · a saved snapshot with a message.
- **Branch** · an independent line of development.
- **Remote** · a copy of the repo hosted elsewhere.
- **Fork** · your personal copy of someone else's repo on GitHub.
- **Pull Request (PR)** · a proposal to merge changes, with review.
- **Clone** · local download of a remote repo.
- **Push / Pull** · send / receive commits to / from a remote.
- **Sync fork** · one-click update of your fork from the original repo.
:::

# The course workflow {background-color="#0067a2"}

## How everything flows

```{mermaid}
flowchart LR
  A[chrglez/<br/>course repo] -->|fork| B[yourname/<br/>course fork]
  B -->|clone| C[RStudio on<br/>your laptop]
  C -->|commit + push| B
  B -->|Pull Request| A
  A -->|review comments| B
  A -.->|Sync fork<br/>web button| B
```

The dashed arrow is the **Sync fork** button on GitHub — that's how
you pull course updates into your fork without a terminal.

## The same story, in words

1. **Fork** the course repo on GitHub — one click, one time only.
2. **Clone** your fork from RStudio.
3. Each day: branch → solve → commit → push.
4. Open a **Pull Request** against `chrglez:main`.
5. Christian reviews with inline comments. PRs are never merged.
6. If the course repo changed while you worked, click **Sync fork** on your fork's page.

## Step 1 · Fork the course repo

Go to <https://github.com/chrglez/quantitative-methods-master-tides>.

Click **Fork** (top-right). GitHub asks for a name — leave the
default. Keep **"Copy the main branch only"** ticked.

You land on your own copy:
`github.com/<your-user>/quantitative-methods-master-tides`.

## Step 2 · Clone your fork from RStudio

On your fork's page, click **Code → HTTPS** and copy the URL.

In RStudio:

1. **File → New Project → Version Control → Git**.
2. Paste the URL into *Repository URL*.
3. Choose a stable parent folder (`~/Documents` or equivalent — **not**
   Desktop or OneDrive).
4. Click **Create Project**.

RStudio clones, opens the project, and the **Git pane** appears in the
top-right.

## Step 3 · Add the `upstream` remote

::: {.callout-note}
This is the **only** terminal command in the whole course.
:::

Inside RStudio, open **Tools → Terminal → New Terminal** and paste:

```bash
git remote add upstream https://github.com/chrglez/quantitative-methods-master-tides.git
git remote -v
```

The second line should print **two** remotes: `origin` (your fork)
and `upstream` (the course repo). After this, close the terminal —
you won't need it again.

## ⚠️ Never commit on `main` {background-color="#fff7ed"}

The `main` branch of your fork is a **read-only mirror** of the
course repo. If you accidentally commit to it, syncing with upstream
becomes painful.

Every day you work on a **new branch**, named after the day:
`day-1`, `day-2`, … `day-6`.

If in doubt, look at the branch selector at the top-right of the Git
pane. It always tells you where you are.

## One branch per day

```{mermaid}
%%| echo: false
gitGraph
  commit id: "course v0"
  commit id: "sync from upstream"
  branch day-1
  checkout day-1
  commit id: "Day 1 solution"
  checkout main
  commit id: "sync from upstream"
  branch day-2
  checkout day-2
  commit id: "Day 2 solution"
  checkout main
  commit id: "sync from upstream"
  branch day-3
  checkout day-3
  commit id: "Day 3 solution"
```

- `main` is a **mirror** of the course repo — it only moves forward
  via *Sync fork*, never by your commits.
- Each `day-N` branch carries **one day's work** and becomes **one PR**.
- Branches are **never merged back** — your fork is the audit trail.

# Day-to-day workflow {background-color="#0067a2"}

## ① Sync your fork · in the browser

Go to `github.com/<your-user>/quantitative-methods-master-tides`.

If Christian updated the course repo while you were away, a banner
appears:

> This branch is N commits behind `chrglez:main` · **Sync fork ▼**

Click it → **Update branch**. Done, one click.

Then in RStudio's Git pane click **Pull** (blue down arrow) to bring
those commits to your laptop.

## ② Start a branch for the day · Git pane

In the Git pane (top-right of RStudio):

1. Click the branch dropdown → **New Branch…**
2. Name it `day-1` (or `day-N`).
3. Leave *Remote: origin* and **tick** *Sync branch with remote*.
4. Click **Create**.

You are now on `day-1`, `main` is safe.

## ③ Solve the exercise

Open `exercises/day1/exercise-template.R`. Fill in your solutions
using what we covered in *Data in R*. **Save** (`Ctrl + S`).

Run the script top-to-bottom (`Ctrl + Shift + Enter`) to confirm it
runs without errors.

## ④ Commit · Git pane

In the Git pane:

1. Tick the checkbox next to `exercises/day1/exercise-template.R`.
2. Click **Commit**.
3. A dialog opens. Write a message: *"Day 1 solution"*.
4. Click **Commit**.

RStudio shows the green confirmation. Your commit exists locally.

## ⑤ Push · Git pane

Click the **green up arrow** (Push) in the Git pane.

RStudio sends your branch to GitHub. The first time per branch it
sets the upstream tracking automatically (because you ticked *Sync
branch with remote* in step ②).

## ⑥ Open the Pull Request · in the browser

Go back to your fork on GitHub. A yellow banner appears:

> **day-1 had recent pushes** · `[ Compare & pull request ]`

Click it. The PR form opens with the right defaults:

- **base** · `chrglez/quantitative-methods-master-tides`, branch `main`
- **compare** · `<your-user>/quantitative-methods-master-tides`, branch `day-1`

Title: `Day 1 — Your Full Name`. Click **Create pull request**.

## That's the whole loop

You have submitted. Christian receives a notification, reads the PR,
leaves inline comments, and eventually closes it.

For the next day, start again from **① Sync your fork**.

## Commit messages that don't suck

::: {.columns}

::: {.column width="50%"}
**Bad**

```text
asdf
update
fix
final version
ok
```
:::

::: {.column width="50%"}
**Good**

```text
Day 1 solution
Add summary table by island
Fix NA handling in occupancy_rate
Rename variable for clarity
```

Imperative mood, starts with a verb, ≤ 50 chars.
:::

:::

## Common pitfalls {.smaller}

::: {.incremental}
- **Never commit secrets** — `.Rdata`, tokens, credentials. The `.gitignore` already excludes most traps.
- **Never commit on `main`** — branches exist for this.
- **First Push** — RStudio handles the upstream tracking automatically if you ticked *Sync branch with remote* when creating the branch.
- **Authentication failed?** Your PAT may have expired. Run `usethis::create_github_token(expiration = 90)` and `gitcreds::gitcreds_set()` again.
- **Large files (>50 MB)** don't belong in Git. Put them under `datasets/raw/` (gitignored) and commit the download script.
:::

## Where to ask for help

- [Happy Git with R](https://happygitwithr.com/) · Jenny Bryan's canonical guide.
- [GitHub Skills](https://skills.github.com/) · short interactive tracks.
- Inside this repo · `exercises/README.md`, tutorial hours on the welcome deck.

# Your turn {background-color="#0067a2"}

In the next 20 minutes, do **all** of these:

1. Fork the course repo and clone it from RStudio.
2. Add `upstream` from the Terminal · one time.
3. Create a `day-1` branch.
4. Write your name at the top of `exercises/day1/exercise-template.R`.
5. Commit and Push from the Git pane.
6. Open the PR on GitHub with the right title.

I'll come around the room to catch anyone stuck.
