---
title: "Git & GitHub essentials"
subtitle: "Module I · Day 1 · 1 hour"
date: "2026-04-23"
---

## Why version control?

::: {.incremental}
- Keep a **full history** of every change — who, what, when, why.
- Collaborate without emailing `analysis_final_v3_reallyfinal.R`.
- **Revert safely** when an experiment breaks your code.
- Make your work **reproducible** and publicly auditable.
:::

::: {.callout-important}
In this course **every exercise you submit lives on GitHub**. The
workflow has to work for you by the end of this hour.
:::

## Git vs GitHub

|  | **Git** | **GitHub** |
|---|---|---|
| What it is | A tool that tracks changes | A web service that hosts Git projects |
| Runs | Locally, on your laptop | On github.com |
| Needed for | Any version control at all | Sharing, collaboration, PRs |
| Cost | Free, open source | Free for public repos and education |

You need **both**: Git installed locally, a GitHub account online.

## The three Git areas

```
┌─────────────┐   git add   ┌─────────────┐  git commit  ┌────────────────┐
│   Working   │ ──────────▶ │   Staging   │ ───────────▶ │  Local repo    │
│  directory  │             │   (index)   │              │  (your laptop) │
└─────────────┘             └─────────────┘              └────────────────┘
                                                                  │
                                                         git push │
                                                                  ▼
                                                         ┌────────────────┐
                                                         │  GitHub remote │
                                                         └────────────────┘
```

Each arrow is a Git command. Mastering the three areas makes the rest fall into place.

## One-off setup · step 1 of 3 {.smaller}

**Install Git**

- macOS · `brew install git`, or install the Xcode Command Line Tools.
- Windows · download [Git for Windows](https://git-scm.com/download/win).
- Linux · `sudo apt install git` / `sudo dnf install git`.

**Configure your identity** (name and email that will sign every commit):

```bash
git config --global user.name  "Your Name"
git config --global user.email "you@ulpgc.es"
git config --global init.defaultBranch main
git config --global pull.rebase false
```

## One-off setup · step 2 of 3 {.smaller}

**Create a GitHub account** at <https://github.com/signup> using your
ULPGC email. Apply for the [Student Pack](https://education.github.com/pack)
— private repos, Copilot and more, free for students.

**Authenticate from R** with a Personal Access Token (PAT):

```{r}
#| echo: true
#| eval: false
install.packages(c("usethis", "gitcreds"))

usethis::create_github_token()   # opens the browser
gitcreds::gitcreds_set()         # paste the token when prompted
```

## One-off setup · step 3 of 3 {.smaller}

**Check everything works** from RStudio:

1. `Tools → Global Options → Git/SVN` — confirm the Git executable is detected.
2. Open this course's `.Rproj` — the *Git* pane should appear in the top-right.
3. From the R console:

```{r}
#| echo: true
#| eval: false
usethis::gh_token_help()   # reports on your PAT status
usethis::git_sitrep()      # full "situation report" on your Git setup
```

If both commands end with green check marks, you are ready.

## Core vocabulary {.smaller}

::: {.incremental}
- **Repository (repo)** · a project folder tracked by Git.
- **Commit** · a saved snapshot with a message.
- **Branch** · an independent line of development.
- **HEAD** · the branch (and commit) you are currently on.
- **Remote** · a copy of the repo hosted elsewhere (e.g. GitHub).
- **Fork** · your personal copy of someone else's repo on GitHub.
- **Pull Request (PR)** · a proposal to merge changes from one branch to another, with review.
- **Clone** · local download of a remote repo.
- **Push / Pull** · send commits to / receive commits from a remote.
:::

# The course workflow {background-color="#0067a2"}

## How everything flows

```{mermaid}
flowchart LR
  A[chrglez/<br/>course repo] -->|fork| B[yourname/<br/>course repo fork]
  B -->|clone| C[your<br/>laptop]
  C -->|commit and<br/>push branch| B
  B -->|Pull Request| A
  A -->|review<br/>comments| B
```

## The same story, in words

1. **Fork** the course repo on GitHub — one click, one time only.
2. **Clone** your fork to your laptop.
3. Solve the day's exercise, commit and push the branch to your fork.
4. Open a **Pull Request** against `chrglez:main`.
5. Christian reviews with inline comments. The PR is the audit trail.

## Branching convention

Each day you work on its own branch, **never on `main`**:

```bash
git checkout main
git pull upstream main       # sync with the course repo
git checkout -b day-1        # start a fresh branch for today
# ...edit files, solve exercise...
git add exercises/day1/exercise-template.R
git commit -m "Day 1 solution"
git push -u origin day-1     # push the new branch to your fork
```

The `-u origin day-1` only needs to be typed the first time you push a
new branch. After that, `git push` is enough.

## Opening the Pull Request

Once you push, GitHub shows a yellow banner on your fork's main page:

> **day-1 had recent pushes** · `[ Compare & pull request ]`

Clicking it opens the PR form:

- **base repository** · `chrglez/quantitative-methods-master-tides` · **base** `main`
- **head repository** · `you/quantitative-methods-master-tides` · **compare** `day-1`
- **Title** · `Day N — Your Full Name`
- **Description** · anything you want Christian to know.

Click **Create pull request** and you are done submitting.

## Live demo · the commands of the day {.smaller}

```bash
# status — what changed?
git status

# diff — what EXACTLY did I change?
git diff

# add — stage specific files (never use -A by default)
git add exercises/day1/exercise-template.R

# commit — snapshot with a descriptive message
git commit -m "Day 1 solution"

# push — upload the branch to your fork
git push

# log — see the history
git log --oneline --graph --decorate
```

Learn these seven and you cover 90 % of what you will ever do.

## Commit messages that don't suck

::: {.columns}

::: {.column width="50%"}
**Bad**

```text
asdf
update
fix
final version
ok
stuff
```

Nobody — not even *you* next week — can tell what those changed.
:::

::: {.column width="50%"}
**Good**

```text
Add summary table by island
Fix NA handling in occupancy_rate
Day 2 solution
Rename variable for clarity
```

Imperative mood, starts with a verb, ≤ 50 chars in the subject line.
:::

:::

## Common pitfalls {.smaller}

::: {.incremental}
- **Never commit secrets** · `.Rdata`, tokens, credentials. The course `.gitignore` already excludes the usual suspects.
- **Never force-push** (`git push --force`) to a shared branch. It rewrites history everyone else pulled.
- If `git push` says *Authentication failed*, regenerate your PAT with `usethis::create_github_token()` and `gitcreds::gitcreds_set()`.
- **Large files** (> 50 MB) do not belong in Git. Put them under `datasets/raw/` (already gitignored) and commit only the download script.
- **Conflicts** during a `pull` are normal, not a disaster. Open the file, look for `<<<<<<<` markers, decide which version wins, `git add` + `git commit`.
:::

## Where to ask for help

- [Happy Git with R](https://happygitwithr.com/) · Jenny Bryan's canonical guide, covers every step of this workflow in depth.
- [Pro Git book](https://git-scm.com/book/en/v2) · free, exhaustive.
- [GitHub Skills](https://skills.github.com/) · short interactive tracks.
- Inside this repo · `exercises/README.md` for the workflow, tutorial hours on the welcome deck.

# Your turn {background-color="#0067a2"}

1. Fork the course repo.
2. Clone your fork.
3. On a `day-1` branch, open `exercises/day1/exercise-template.R` and add your name in the header.
4. `git add` → `git commit -m "Day 1 · hello"` → `git push -u origin day-1`.
5. Open the Pull Request on GitHub.

Show me the PR URL in class.
