---
title: "Data wrangling with the tidyverse"
subtitle: "Topic 1 · Day 2 · 4 hours"
date: last-modified
---

## Branch *before* you edit {background-color="#fff7ed"}

Yesterday a few of you committed straight to `main`. Easy to fix, but
let's get the muscle memory right today.

**The two clicks, every day, before opening the exercise template:**

1. **Pull** · Git pane → blue down arrow (after *Sync fork* on GitHub).
2. **New Branch** · Git pane → branch dropdown → *New Branch…* → name
   `day-2` → tick *Sync branch with remote* → *Create*.

Look at the branch selector at the top of the Git pane. If it does
**not** say `day-2`, stop and fix it before you type anything in the
template.

::: {.callout-tip}
Mantra: *Sync · Pull · Branch · then* edit.
:::

## Outline

::: {.incremental}
- The pipe `|>` and the tidyverse grammar.
- `dplyr` verbs: `filter`, `select`, `mutate`, `arrange`, `summarise`, `group_by`.
- Reshaping with `tidyr`: `pivot_longer`, `pivot_wider`, `separate`, `unite`.
- Joining tables: `left_join`, `inner_join`, `anti_join`.
- Missing values and type coercion gotchas.
:::

## The pipe

```{r}
#| echo: true
#| eval: false
hotels |>
  filter(nights > 7) |>
  mutate(revenue = nights * price) |>
  arrange(desc(revenue))
```

Read as: *take `hotels`, keep long stays, compute revenue, sort
descending*.

## The six `dplyr` verbs

| Verb | Purpose |
|---|---|
| `filter()`   | keep rows that match a condition |
| `select()`   | keep / drop columns |
| `mutate()`   | add or transform columns |
| `arrange()`  | sort rows |
| `summarise()`| collapse to one row per group |
| `group_by()` | set the grouping context |

## Group-wise summaries · Eurostat nights spent

```{r}
#| echo: true
#| eval: false
library(dplyr)

nights |>
  filter(year == 2024, purpose == "total") |>
  group_by(country) |>
  summarise(
    total_nights = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_nights))
```

## Wide vs long · `pivot_longer()`

```{r}
#| echo: true
#| eval: false
library(tidyr)

nights_wide <- nights |>
  pivot_wider(names_from = year, values_from = value)

nights_wide |>
  pivot_longer(cols = `2019`:`2024`,
               names_to = "year", values_to = "value")
```

## Joins · occupancy + capacity

```{r}
#| echo: true
#| eval: false
occupancy |>
  left_join(capacity, by = c("island", "year", "month")) |>
  mutate(occupancy_rate = nights / available_beds)
```

Know your joins:

- `left_join()` — keep every row of the left table.
- `inner_join()` — keep only matches.
- `anti_join()` — rows from left **without** a match on the right.

## NA handling

```{r}
#| echo: true
x <- c(2, NA, 7, 4)
mean(x)                # NA
mean(x, na.rm = TRUE)  # 4.33
sum(is.na(x))          # count NAs
```

## Recap

- Pipe `|>` chains verbs, reads left-to-right.
- `group_by() |> summarise()` is the bread and butter of reporting.
- Always pivot when a column stores values, not variables.

# Hands-on {background-color="#0067a2"}

Join an ISTAC occupancy file with an ISTAC capacity file, compute
monthly occupancy rate by island, and produce a summary table for 2024.
