---
title: "Basic statistics with R"
subtitle: "Topic 1 · Day 3 · 2 hours"
date: last-modified
---

## Outline

::: {.incremental}
- Univariate descriptives: mean, median, variance, quantiles.
- Frequency tables and cross-tabulations.
- Correlation and covariance.
- Communicating results with nicely formatted summaries.
:::

::: {.callout-note appearance="simple"}
**Hypothesis testing** (t-test, chi-squared, regression inference)
is covered in **Module III** with Juan — not here.
:::

## Setup · sample data

Run this once at the start of the session — every example below uses
either `hotels` or `tourists`.

```{r}
#| echo: true
#| eval: false
library(tibble)

hotels <- tibble(
  island = c("Gran Canaria", "Tenerife", "Lanzarote",
             "Fuerteventura", "La Palma"),
  stars  = c(4L, 5L, 4L, 3L, 3L),
  price  = c(82, 95, 110, 100, 78),
  rating = c(8.2, 9.1, 7.5, 6.9, 8.0),
  nights = c(12.5, 18.3, 9.8, 11.2, 6.4)
)

tourists <- tibble(
  origin_country = c("DE", "DE", "UK", "UK", "UK",
                     "ES", "ES", "FR", "DE", "UK"),
  island         = c("Gran Canaria", "Tenerife", "Lanzarote",
                     "Tenerife", "Lanzarote", "Gran Canaria",
                     "Gran Canaria", "Tenerife", "Gran Canaria",
                     "Lanzarote"),
  main_purpose   = c("leisure", "leisure", "leisure", "business",
                     "leisure", "leisure", "business", "leisure",
                     "leisure", "leisure")
)
```

## Descriptives in one line

```{r}
#| echo: true
#| eval: false
library(dplyr)

hotels |>
  summarise(
    n        = n(),
    mean     = mean(price, na.rm = TRUE),
    sd       = sd(price,   na.rm = TRUE),
    median   = median(price, na.rm = TRUE),
    q25      = quantile(price, 0.25, na.rm = TRUE),
    q75      = quantile(price, 0.75, na.rm = TRUE)
  )
```

## Frequency tables

```{r}
#| echo: true
#| eval: false
library(janitor)

tourists |>
  tabyl(origin_country) |>
  adorn_totals("row") |>
  adorn_pct_formatting()
```

## Cross-tabs

```{r}
#| echo: true
#| eval: false
tourists |>
  tabyl(island, main_purpose) |>
  adorn_percentages("row") |>
  adorn_pct_formatting()
```

## Correlation

```{r}
#| echo: true
#| eval: false
hotels |>
  select(price, rating, nights) |>
  cor(use = "pairwise.complete.obs")
```

## Recap

- `summarise()` + a handful of functions covers most univariate reporting.
- `janitor::tabyl()` beats base `table()` for presentable frequency tables.
- `cor()` with `use = "pairwise.complete.obs"` handles missing values
  sensibly when summarising relationships among numeric variables.

# Hands-on {background-color="#0067a2"}

Using the `hotels` and `tourists` tibbles from the Setup slide:

1. Summary table of mean price, mean rating and mean nights, grouped
   by `stars`.
2. Cross-tabulation of `origin_country` × `island` with row
   percentages, and one sentence on a striking cell.
3. Correlation matrix of `price`, `rating` and `nights` with one
   sentence on the strongest pair.

The graded version with real Eurostat data is in
`exercises/day3/exercise-template.R`.
