---
title: "Basic statistics with R"
subtitle: "Topic 1 · Day 3 · 2 hours"
date: last-modified
---

## Outline

::: {.incremental}
- Univariate descriptives: mean, median, variance, quantiles.
- Frequency tables and cross-tabulations.
- Correlation and covariance.
- Inference 101: t-test, chi-squared, proportion test.
- Communicating results: `broom::tidy()` and nicely formatted tables.
:::

## Descriptives in one line

```{r}
#| echo: true
#| eval: false
library(dplyr)

hotels |>
  summarise(
    n        = n(),
    mean     = mean(price, na.rm = TRUE),
    sd       = sd(price,   na.rm = TRUE),
    median   = median(price, na.rm = TRUE),
    q25      = quantile(price, 0.25, na.rm = TRUE),
    q75      = quantile(price, 0.75, na.rm = TRUE)
  )
```

## Frequency tables

```{r}
#| echo: true
#| eval: false
library(janitor)

tourists |>
  tabyl(origin_country) |>
  adorn_totals("row") |>
  adorn_pct_formatting()
```

## Cross-tabs

```{r}
#| echo: true
#| eval: false
tourists |>
  tabyl(island, main_purpose) |>
  adorn_percentages("row") |>
  adorn_pct_formatting()
```

## Correlation

```{r}
#| echo: true
#| eval: false
hotels |>
  select(price, rating, nights) |>
  cor(use = "pairwise.complete.obs")
```

## A simple t-test · does price differ between two islands?

```{r}
#| echo: true
#| eval: false
t.test(price ~ island,
       data   = filter(hotels, island %in% c("Gran Canaria", "Tenerife")))
```

## Tidy inference with `broom`

```{r}
#| echo: true
#| eval: false
library(broom)

t.test(price ~ island,
       data = filter(hotels, island %in% c("Gran Canaria", "Tenerife"))) |>
  tidy()
```

## Recap

- `summarise()` + a handful of functions covers most univariate reporting.
- `janitor::tabyl()` beats base `table()` for presentable frequency tables.
- `broom::tidy()` converts any test into a one-row data frame.

# Hands-on {background-color="#0067a2"}

Produce a summary table of average nightly price and average stay
length by island, and test whether nights spent in 2024 differ between
two Canary islands of your choice.
