---
title: "Day 6 · Final submission"
subtitle: "A reproducible report on Canary Islands tourism"
author: "Ilona Maier"
date: today
format:
  html:
    toc: true
  pdf:
    toc: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(here)
library(gt)
theme_set(theme_minimal(base_size = 11))
```

# 1. Research question

How does tourist spending and length of stay vary across the Canary Islands?
This is relevant because tourism is the main economic activity of the
Canary Islands. In the context of overtourism, islands have to try to achieve
higher economic output with lower numbers of tourists. The higher the 
spending per night, the better. Also it is more sustainable if less tourists
come that stay longer, because there are less in and outbound flights.

# 2. Data

The data comes from the ISTAC Tourist Expenditure Survey 2025, a 
quarterly survey conducted at Canary Islands airports. 
It covers tourist departures in 2025 and includes information on spending, 
length of stay, island visited, nationality, and purpose of visit. The dataset 
contains 9,083 individual responses.

For this analysis, I selected five variables: island, nights, total spending, 
purpose of visit, and nationality. I removed respondents with zero nights or 
zero spending, and excluded records with unidentified island codes. I also 
computed a new variable, spending per night, by dividing total spending by 
number of nights. The spending includes accommodation and other spending, but
does not include the spending for flights.

```{r}
# your code here

# Load data
dat <- read_csv(
  here("datasets", "raw", "gastoturistico2025.csv"),
  locale = locale(encoding = "latin1")
)

# Clean and select relevant variables
tourists <- dat |>
  select(
    island    = ISLA,
    nights    = NOCHES,
    spending  = GASTO_EUROS,
    accom     = COSTE_ALOJ_EUROS,
    purpose   = PROPOSITO,
    nationality = NACIONALIDAD
  ) |>
  mutate(
    spending_per_night = (spending + accom) / nights,
    island = case_when(
      island == "ES705" ~ "Gran Canaria",
      island == "ES709" ~ "Tenerife",
      island == "ES708" ~ "Lanzarote",
      island == "ES704" ~ "Fuerteventura",
      island == "ES707" ~ "La Palma",
      island == "ES706" ~ "La Gomera",
      island == "ES703" ~ "El Hierro",
      TRUE ~ NA_character_
   ) )|>
  filter(
    nights > 0,
    spending > 0,
    !is.na(island)  )

# Summary table with gt
summary_table <- tourists |>
  group_by(island) |>
  summarise(
    n           = n(),
    mean_nights  = round(mean(nights, na.rm = TRUE), 1),
    mean_spn = round(mean(spending_per_night, na.rm = TRUE), 0)  ) |>
  arrange(desc(n)) |>
  gt() |>
  cols_label(
    island        = "Island",
    n             = "Tourists",
    mean_nights   = "Avg. Nights",
    mean_spn = "Avg. Spending/Night (€)"
  ) |>
  tab_header(
    title    = "Tourist stays and spending by island",
    subtitle = "Canary Islands · 2025 · ISTAC"
  ) |>
  tab_source_note("Source: ISTAC, Encuesta de Gasto Turístico 2025.")

summary_table
```

```

```

# 3. Descriptive analysis

```{r}
# your code here

# ---- Plot 1: Univariate - Distribution of spending per night ----------
plot1 <- ggplot(tourists, aes(x = spending_per_night)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "#0067a2", alpha = 0.55) +
  geom_density(linewidth = 1) +
  xlim(0,1000)+
  labs(
    title   = "Distribution of Spending per Night",
    x       = "Spending per night (€)",
    y       = "Density",
    caption = "Source: ISTAC, Encuesta de Gasto Turístico 2025"
  )

plot1

# ---- Plot 2: Spending per night by island -----------------------------
plot2 <- ggplot(tourists, aes(x = reorder(island, spending_per_night,
                                           median), y = spending_per_night)) +
  geom_boxplot(fill = "#0067a2", colour = "black") +
  coord_flip() +
  coord_cartesian(ylim = c(0, 750)) +
  labs(
    title   = "Spending per Night by Island",
    x       = NULL,
    y       = "Spending per night (€)",
    caption = "Source: ISTAC, Encuesta de Gasto Turístico 2025"
  )

plot2

# ---- Plot 3: Length of stay by island ---------------------------------
plot3 <- ggplot(tourists, aes(x = reorder(island, nights, median),
                               y = nights)) +
  geom_boxplot(fill = "#f4f3ee", colour = "#0067a2") +
  coord_cartesian(ylim = c(0, 30)) +
  labs(
    title   = "Length of Stay by Island",
    x       = NULL,
    y       = "Nights",
    caption = "Source: ISTAC, Encuesta de Gasto Turístico 2025"
  )

plot3

# ---- Inferential test: ANOVA ------------------------------------------
# Test whether spending per night differs significantly across islands
anova_result <- aov(spending_per_night ~ island, data = tourists)

# Show ANOVA result as clean table
anova_result |>
  broom::tidy() |>
  gt() |>
  fmt_number(columns = c(sumsq, meansq, statistic), decimals = 2) |>
  fmt_number(columns = p.value, decimals = 4) |>
  cols_label(
    term      = "Source",
    df        = "df",
    sumsq     = "Sum of Squares",
    meansq    = "Mean Square",
    statistic = "F value",
    p.value   = "p-value"
  ) |>
  tab_header(
    title    = "ANOVA: Spending per Night by Island",
    subtitle = "Source: ISTAC, Encuesta de Gasto Turístico 2025"
  )

```

```

# 4. Conclusions

Plot 1 shows that spending per night is right-skewed, meaning most 
tourists spend between €50 and €250 per night, but a small number 
spend considerably more.

Plot 2 shows clear differences between islands. Lanzarote has the 
highest median spending per night, followed closely by Tenerife. 
La Palma has the lowest median. Gran Canaria and Fuerteventura fall in between.
All islands show wide variation, suggesting large differences between tourist 
profiles even within the same island.

Plot 3 shows that length of stay is remarkably similar across all islands, 
with a median of 7 nights for every island. This suggests that the typical 
tourist stays exactly one week regardless of which island they visit, likely 
reflecting standard package holiday durations. Gran Canaria and La Palma show 
slightly more variation, with some tourists staying up to 9-10 nights (Q75). 
Tenerife and Lanzarote have the most concentrated stays, suggesting a more 
standardised tourism product.

A one-way ANOVA was conducted to test whether average spending per night differs significantly across islands. The result was statistically significant (F = 4.38, p = 0.0015), meaning the differences in spending per night between islands are unlikely to be due to chance alone. This supports the visual finding from Plot 2 that Lanzarote and Tenerife have higher spending per night than La Palma. However, the effect size appears modest given the large sample size (n = 8,726), and the result should be interpreted with caution.

One limitation is that the data only captures tourists surveyed at airports 
upon departure, which means tourists who left by other means (e.g. ferry) 
are excluded. This could introduce a selection bias, particularly for smaller 
islands like La Palma where ferry travel is more common. Also El Hierro and 
La Gomera are not included in the analysis.


# 5. Reproducibility checklist

- [x] The `.qmd` renders end-to-end from a clean clone.
- [x] Data files live under `datasets/raw/` (gitignored) and are
  downloaded by a script in `datasets/download.R`.
- [x] No absolute paths. All paths go through `here::here()`.
- [x] The repository is a fork of the course repo, pushed to
  `github.com/<you>/quantitative-methods-master-tides`.
