---
title: "Day 6 · Final submission"
subtitle: "A reproducible report on Canary Islands tourism"
author: "Your Name"
date: today
format:
  html:
    toc: true
  pdf:
    toc: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(here)
library(gt)
theme_set(theme_minimal(base_size = 11))
```

# 1. Research question

State in one paragraph the tourism question you address and why it
matters.

# 2. Data

Describe the data source (ISTAC or Eurostat), the time coverage, and
the cleaning steps you applied. Include one summary table produced
with `gt`.

```{r}
# your code here
```

# 3. Descriptive analysis

Include at least three plots (one univariate, two multivariate) and
one inferential test. Every plot must have a title, labelled axes, and
a source line.

```{r}
# your code here
```

# 4. Conclusions

Two-paragraph conclusion. State what you found, what you did **not**
find, and one limitation of your analysis.

# 5. Reproducibility checklist

- [ ] The `.qmd` renders end-to-end from a clean clone.
- [ ] Data files live under `datasets/raw/` (gitignored) and are
  downloaded by a script in `datasets/download.R`.
- [ ] No absolute paths. All paths go through `here::here()`.
- [ ] The repository is a fork of the course repo, pushed to
  `github.com/<you>/quantitative-methods-master-tides`.
