---
title: "Day 1 · Exercise"
subtitle: "Git, GitHub and Data in R"
author: "Your Name"
date: today
format: html
---

```{r}
#| include: false
library(tidyverse)
library(here)
```

# 1. Identify yourself

Replace the placeholder with your name and GitHub handle.

- **Name:** Your Name
- **GitHub:** @yourhandle

# 2. R as a calculator

Compute the sum of the first 100 positive integers and the geometric
mean of the vector `c(2, 4, 8, 16)`. Write both results below.

```{r}
# your code here
```

# 3. Vectors

Build a vector of the monthly average temperatures in Las Palmas
(°C) from a source of your choice. Compute its minimum, maximum,
mean, and standard deviation.

```{r}
# your code here
```

# 4. Load a tourism dataset

Load one of the ISTAC CSV files placed under `datasets/raw/`. Use
`glimpse()` to inspect it and report in prose:

- how many rows and columns it has,
- the types of each column,
- whether there are missing values.

```{r}
# your code here
```

# 5. Commit and push

Once you are happy with this document, run in a terminal:

```bash
git add exercises/day1/exercise-template.qmd
git commit -m "Day 1 solution — <your name>"
git push
```

Then paste the URL of the committed file on your fork in the course
submission form.
