---
title: "One-variable visualisation"
subtitle: "Topic 2 · Day 4 · 2 hours"
date: last-modified
---

## Outline

::: {.incremental}
- Why visualise before you model.
- Distribution at a glance: histograms, density, rug plots.
- Robust alternatives: boxplots and violin plots.
- Categorical: bars, lollipop, waffle.
- From base R to `ggplot2`: the grammar of graphics.
:::

## Base R · quick and local

```{r}
#| echo: true
#| eval: false
hist(hotels$price,
     breaks = 20,
     col    = "#0067a2",
     main   = "Distribution of nightly price",
     xlab   = "Price (€)")
```

## `ggplot2` · the grammar

```{r}
#| echo: true
#| eval: false
library(ggplot2)

ggplot(hotels, aes(x = price)) +
  geom_histogram(bins = 30, fill = "#0067a2", alpha = 0.85) +
  labs(title = "Distribution of nightly price",
       x = "Price (€)", y = "Hotels") +
  theme_minimal(base_size = 12)
```

## Histogram vs density

```{r}
#| echo: true
#| eval: false
ggplot(hotels, aes(x = price)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "#0067a2", alpha = 0.55) +
  geom_density(linewidth = 1) +
  theme_minimal()
```

## Boxplot · robust summary

```{r}
#| echo: true
#| eval: false
ggplot(hotels, aes(y = price)) +
  geom_boxplot(fill = "#f4f3ee") +
  coord_flip() +
  theme_minimal()
```

## Categorical · ordered bar chart

```{r}
#| echo: true
#| eval: false
library(forcats)

tourists |>
  count(origin_country) |>
  mutate(origin_country = fct_reorder(origin_country, n)) |>
  ggplot(aes(n, origin_country)) +
  geom_col(fill = "#0067a2") +
  labs(x = "Tourists", y = NULL) +
  theme_minimal()
```

## Rules of thumb

- **One variable → one plot type**: continuous → histogram/density;
  categorical → bar chart.
- Order matters for categories — reorder by frequency or by value.
- Minimise ink; use `theme_minimal()` (or `theme_void()` for maps).
- Titles, units, source line — always.

# Hands-on {background-color="#0067a2"}

Produce publication-ready histogram and boxplot of **nights spent**
in an ISTAC series of your choice, for 2024. Push to your fork.
