---
title: "Multi-variable visualisation"
subtitle: "Topic 2 · Day 5 · 2 hours"
date: last-modified
---

## Outline

::: {.incremental}
- Mapping aesthetics: `x`, `y`, `colour`, `size`, `shape`, `alpha`.
- Two continuous variables: scatter, with smoothers.
- One continuous + one categorical: side-by-side distributions.
- Facets: small multiples.
- Correlation matrices and heatmaps.
- Mapping tourism data to Canary Islands geography.
:::

## Scatter · price vs rating

```{r}
#| echo: true
#| eval: false
library(ggplot2)

ggplot(hotels, aes(x = rating, y = price, colour = island)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Rating (1–10)", y = "Nightly price (€)") +
  theme_minimal()
```

## Facets · one panel per island

```{r}
#| echo: true
#| eval: false
ggplot(hotels, aes(x = nights, y = price)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ island, ncol = 4) +
  theme_minimal()
```

## Categorical × continuous · violin + boxplot

```{r}
#| echo: true
#| eval: false
ggplot(hotels, aes(x = island, y = price)) +
  geom_violin(fill = "#f4f3ee", linewidth = 0.3) +
  geom_boxplot(width = 0.2, fill = "#0067a2", colour = "white") +
  coord_flip() +
  theme_minimal()
```

## Correlation heatmap

```{r}
#| echo: true
#| eval: false
library(dplyr)
library(ggplot2)
library(tidyr)

hotels |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs") |>
  as.data.frame() |>
  tibble::rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile() +
  scale_fill_gradient2(midpoint = 0, limits = c(-1, 1)) +
  theme_minimal()
```

## Maps · ISTAC data on Canary Islands

```{r}
#| echo: true
#| eval: false
library(sf)
library(ggplot2)

# Shape file for Canary Islands municipalities (download separately).
muni <- read_sf("datasets/raw/canarias-municipios.geojson")

muni |>
  left_join(occupancy_by_muni, by = "muni_code") |>
  ggplot() +
  geom_sf(aes(fill = nights_per_capita), colour = "white", linewidth = 0.2) +
  scale_fill_viridis_c() +
  theme_void()
```

## Principles

- **One story per plot.** If it takes a paragraph to read, split it.
- Colour should encode **one** variable at a time.
- Prefer facets over a forest of overlaid lines.
- Check for overplotting — use `alpha`, jitter, or 2D binning.

# Hands-on {background-color="#0067a2"}

Produce a faceted scatterplot of Eurostat *nights spent* vs *GDP per
capita* by country, faceted by year, from 2019 to 2024.
