---
title: "Multi-variable visualisation"
subtitle: "Topic 2 · Day 5 · 2 hours"
date: last-modified
---

## Outline

::: {.incremental}
- Mapping aesthetics: `x`, `y`, `colour`, `size`, `shape`, `alpha`.
- Two continuous variables: scatter, with smoothers.
- One continuous + one categorical: side-by-side distributions.
- Facets: small multiples.
- Correlation matrices and heatmaps.
:::

## Setup · sample data

Run this once at the start of the session — every example below uses
`hotels`.

```{r}
#| echo: true
#| eval: false
library(tibble)

hotels <- tibble(
  island = c("Gran Canaria", "Tenerife", "Lanzarote",
             "Fuerteventura", "La Palma"),
  stars  = c(4L, 5L, 4L, 3L, 3L),
  price  = c(82, 95, 110, 100, 78),
  rating = c(8.2, 9.1, 7.5, 6.9, 8.0),
  nights = c(12.5, 18.3, 9.8, 11.2, 6.4)
)
```

## Scatter · price vs rating

```{r}
#| echo: true
#| eval: false
library(ggplot2)

ggplot(hotels, aes(x = rating, y = price, colour = island)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Rating (1–10)", y = "Nightly price (€)") +
  theme_minimal()
```

## Facets · one panel per island

```{r}
#| echo: true
#| eval: false
ggplot(hotels, aes(x = nights, y = price)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ island, ncol = 4) +
  theme_minimal()
```

## Categorical × continuous · violin + boxplot

```{r}
#| echo: true
#| eval: false
ggplot(hotels, aes(x = island, y = price)) +
  geom_violin(fill = "#f4f3ee", linewidth = 0.3) +
  geom_boxplot(width = 0.2, fill = "#0067a2", colour = "white") +
  coord_flip() +
  theme_minimal()
```

## Correlation heatmap

```{r}
#| echo: true
#| eval: false
library(dplyr)
library(ggplot2)
library(tidyr)

hotels |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs") |>
  as.data.frame() |>
  tibble::rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile() +
  scale_fill_gradient2(midpoint = 0, limits = c(-1, 1)) +
  theme_minimal()
```

## Principles

- **One story per plot.** If it takes a paragraph to read, split it.
- Colour should encode **one** variable at a time.
- Prefer facets over a forest of overlaid lines.
- Check for overplotting — use `alpha`, jitter, or 2D binning.

# Hands-on {background-color="#0067a2"}

Using the Eurostat monthly nights file already downloaded
(`datasets/raw/eurostat-nights_monthly.csv`), pick 4-6 EU countries
and produce a faceted line plot:

- one panel per country (`facet_wrap(~ geo)`),
- x = month, y = monthly nights,
- one line per year (`colour = factor(year)`),
- 2019-2024.

The plot should make seasonality and the COVID dip visible at a
glance.
