---
title: "Day 3 · Exercise (graded)"
subtitle: "Basic statistics with R"
author: "Your Name"
date: today
format: html
---

```{r}
#| include: false
library(tidyverse)
library(janitor)
library(broom)
library(here)
```

# 1. Descriptives

Using ISTAC or Eurostat data, produce a summary table (n, mean, SD,
median, Q1, Q3, missing) for two continuous tourism variables of your
choice, broken down by a categorical grouping variable (e.g. island,
country, quarter).

```{r}
# your code here
```

# 2. Cross-tabulation

Build one cross-tabulation with row percentages. Briefly comment on
any striking cell.

```{r}
# your code here
```

# 3. A comparison test

State a hypothesis of the form "mean of Y in group A equals mean of Y
in group B" and test it with a two-sample t-test. Report the point
estimates, confidence interval, and p-value using `broom::tidy()`.

```{r}
# your code here
```

# 4. One paragraph of interpretation

Interpret the test in plain English for a non-statistical tourism
reader. Say what "statistically significant" means and what it does
**not** mean.
